---
title: "Data608 Module 3 Assignment"
author: "Chinedu Onyeka"
date: '2022-10-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the required libraries
```{r load-lib, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(shiny)
```

Read in the data
```{r read-data,message=FALSE, warning=FALSE}
cdc_mortality_df <- read.csv('https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module3/data/cleaned-cdc-mortality-1999-2010-2.csv')
cdc_mortality_df <- as_tibble(cdc_mortality_df)
```

Check the head
```{r}
head(cdc_mortality_df)
```
Check the summary
```{r}
summary(cdc_mortality_df)
```

```{r rename_col, warning=FALSE}
# rename the ICD.Chapter to ICD_chapter and Crude.Rate to Crude_Rate
cdc_mortality_df <- cdc_mortality_df %>% rename(ICD_Chapter = ICD.Chapter, Crude_Rate = Crude.Rate)
cdc_mortality_df
```
```{r}
df = cdc_mortality_df
states <- unique(df$State)

df_summary <-  df %>% 
               group_by(Year,ICD_Chapter) %>% 
               summarise(State = "Average",
                    Deaths = mean(Deaths),
                    Population = mean(Population),
                    Crude_Rate = 100000 * sum(Deaths)/sum(Population)) %>% 
               bind_rows(df)
df2 <- df_summary
```


#### Question 1
As a researcher, you frequently compare mortality rates from particular causes across different States. You need a visualization that will let you see (for 2010 only) the crude mortality rate, across all States, from one cause (for example, Neoplasms, which are effectively cancers). Create a visualization that allows you to rank States by crude mortality for each cause of death.  

```{r filter2010}
#Filter the data for only 2010 and from one cause say Neoplasms
cdc_mortality_df_2010 = cdc_mortality_df %>% filter(Year == 2010 & ICD_Chapter == 'Neoplasms') %>% 
                        arrange(State, desc(Crude_Rate))
head(cdc_mortality_df_2010)
```
ShinyApp for Question1

```{r shinyApp1}

# Build the UI for question1
ui <- fluidPage(
  # Panel for Question 1
  tabsetPanel( 
    tabPanel(title = "Question1",
      sidebarPanel(
        htmlOutput('message_q1'),
        sliderInput('year_q1','Year',min(df$Year),max(df$Year),2010,1,width=600,
                  sep = ""),
        selectInput('causes_q1', 'Causes', 
                  unique(df$ICD_Chapter), selected='Neoplasms',width = 600)
      ),
      mainPanel(plotlyOutput('plot1_q1'))
    )
    ))

# build the server for question1
server <- function(input, output, session) {
  # Question1 Server Side
  data_q1 <- reactive({
    df2 <- df %>%
       filter(Year == input$year_q1) %>%
       filter(ICD_Chapter == input$causes_q1) %>%
       arrange(Crude_Rate) 
    df2$State <- factor(df2$State, levels = unique(df2$State))
    df2
  })
  # output message for question1
  output$message_q1 <- renderText({
    
    df2 <- data_q1()
    if (nrow(df2) == 0) {
      paste("<h4><center>No Data for<b>",input$causes_q1,
            "</b>in",input$year_q1,"</center></h4>", sep =" ") 
    }
    else {
      paste("<h4><center>The Mortality Rates for<b>",input$causes_q1,
            "</b>in",input$year_q1,"<br>(scaled by multiplying 100000)</center></h4>", sep =" ")
    }
  })
  
  # output plot for question1
  output$plot1_q1 <- renderPlotly({

    df2 <- data_q1()
    if (nrow(df2) == 0) {
      plotly_empty(type = "bar",width = 1,height = 1,)
    }
    else {
    plot_ly(x = df2$Crude_Rate,
            y = df2$State,
            orientation='h',
            width = 900,
            height = 750,
            type = "bar") %>% 
        layout(
          yaxis = list(tickfont = list(size = 11)))
    }
  }
  )
}

shinyApp(ui = ui, server = server)
```


#### Question 2
Often you are asked whether particular States are improving their mortality rates (per cause) faster than, or slower than, the national average. Create a visualization that lets your clients see this for themselves for one cause of death at the time. Keep in mind that the national average should be weighted by the national population.

```{r shinyApp2}
# Build the UI part of the shinyApp
ui <- fluidPage(
  tabsetPanel( 
    # Panel for Question 2
    tabPanel(title = "Question2",
      sidebarPanel(
        htmlOutput('message_q2'),
        selectInput('causes_q2', 'Causes', 
                  unique(df$ICD_Chapter), selected='Neoplasms',width = 600),
        checkboxGroupInput("States_q2", "States",states,inline = TRUE)
      ),
      mainPanel(plotlyOutput('plot1_q2'))
    )
  )
)

# Build the server side of the shinyApp

server <- function(input, output, session) {
  #Question2 Server Side
  data_q2 <- reactive({
    df3 <- df2 %>%
      filter(ICD_Chapter == input$causes_q2) 
    df3
  })
  
  state2_q2 <-reactive(
    input$States_q2
  )
  
  # output message for question2
  output$message_q2 <- renderText({
    paste("<h4><center>The Mortality Rates for<b>",input$causes_q2,
                   "</b><br>(scaled by multiplying 100000) </center></h4>", sep =" ")
    
  })
  
  # output plot for question2
  output$plot1_q2 <- renderPlotly({
    
    df3 <- data_q2()
    if (nrow(df3) == 0) {
      plotly_empty(type = "scatter",width = 1,height = 1,)
    }
    else {
      temp <- filter(df3, State == "Average")
      
      fig <- plot_ly(
        x = temp$Year,
        y = temp$Crude_Rate,
        mode = 'lines+markers',
        name = "Average",
        type = "scatter"
      )
      selected_states <- state2_q2()
      for (state in states)
      {
        alpha <- ifelse(state %in% selected_states,1,0.05)
        temp <- filter(df3, State == state)
        fig <- fig %>% add_trace(x = temp$Year, y = temp$Crude_Rate, name = state, mode = 'lines+markers',opacity = alpha) 
      }
      fig
    }
  })
}

shinyApp(ui = ui, server = server)
```

### Combined ShinyApp for question1 and question2

```{r shinyApp}

# Build the UI part of the shinyApp
ui <- fluidPage(
  # Panel for Question 1
  tabsetPanel( 
    tabPanel(title = "Question1",
      sidebarPanel(
        htmlOutput('message_q1'),
        sliderInput('year_q1','Year',min(df$Year),max(df$Year),2010,1,width=600,
                  sep = ""),
        selectInput('causes_q1', 'Causes', 
                  unique(df$ICD_Chapter), selected='Neoplasms',width = 600)
      ),
      mainPanel(plotlyOutput('plot1_q1'))
    ),
    #---------------------------------------------------------------------------
    
    # Panel for Question 2
    tabPanel(title = "Question2",
      sidebarPanel(
        htmlOutput('message_q2'),
        selectInput('causes_q2', 'Causes', 
                  unique(df$ICD_Chapter), selected='Neoplasms',width = 600),
        checkboxGroupInput("States_q2", "States",states,inline = TRUE)
      ),
      mainPanel(plotlyOutput('plot1_q2'))
    )
  )
)

# Build the server side of the shinyApp

server <- function(input, output, session) {
  # Question1 Server Side
  data_q1 <- reactive({
    df2 <- df %>%
       filter(Year == input$year_q1) %>%
       filter(ICD_Chapter == input$causes_q1) %>%
       arrange(Crude_Rate) 
    df2$State <- factor(df2$State, levels = unique(df2$State))
    df2
  })
  # output message for question1
  output$message_q1 <- renderText({
    
    df2 <- data_q1()
    if (nrow(df2) == 0) {
      paste("<h4><center>No Data for<b>",input$causes_q1,
            "</b>in",input$year_q1,"</center></h4>", sep =" ") 
    }
    else {
      paste("<h4><center>The Mortality Rates for<b>",input$causes_q1,
            "</b>in",input$year_q1,"<br>(scaled by multiplying 100000)</center></h4>", sep =" ")
    }
  })
  
  # output plot for question1
  output$plot1_q1 <- renderPlotly({

    df2 <- data_q1()
    if (nrow(df2) == 0) {
      plotly_empty(type = "bar",width = 1,height = 1,)
    }
    else {
    plot_ly(x = df2$Crude_Rate,
            y = df2$State,
            orientation='h',
            width = 900,
            height = 750,
            type = "bar") %>% 
        layout(
          yaxis = list(tickfont = list(size = 11)))
    }
  })
  
#-------------------------------------------------------------------------------
  #Question2 Server Side
  data_q2 <- reactive({
    df3 <- df2 %>%
      filter(ICD_Chapter == input$causes_q2) 
    df3
  })
  
  state2_q2 <-reactive(
    input$States_q2
  )
  
  # output message for question2
  output$message_q2 <- renderText({
    paste("<h4><center>The Mortality Rates for<b>",input$causes_q2,
                   "</b><br>(scaled by multiplying 100000) </center></h4>", sep =" ")
    
  })
  
  # output plot for question2
  output$plot1_q2 <- renderPlotly({
    
    df3 <- data_q2()
    if (nrow(df3) == 0) {
      plotly_empty(type = "scatter",width = 1,height = 1,)
    }
    else {
      temp <- filter(df3, State == "Average")
      
      fig <- plot_ly(
        x = temp$Year,
        y = temp$Crude_Rate,
        mode = 'lines+markers',
        name = "Average",
        type = "scatter"
      )
      selected_states <- state2_q2()
      for (state in states)
      {
        alpha <- ifelse(state %in% selected_states,1,0.05)
        temp <- filter(df3, State == state)
        fig <- fig %>% add_trace(x = temp$Year, y = temp$Crude_Rate, name = state, mode = 'lines+markers',opacity = alpha) 
      }
      fig
    }
  })
}

shinyApp(ui = ui, server = server)
```






